---
title: 'Security incidents in Mali: the incidents'
author: Fousseynou Bah
date: '2018-05-15'
categories:
  - data
  - R
tags:
  - data
  - Mali
  - R Markdown
  - security
  - terrorism
  - threats
slug: security-incidents-in-mali-the-incidents
---


```{r setup, include=FALSE}
#knitr::opts_chunk$set(echo = FALSE)
Sys.setenv(TZ="Africa/Bamako")
##########################################################################
########################################/// PACKAGES ///
##########################################################################
library("tidyverse")
#library("ggvis")
library("lubridate")
library("plotly")
#library("readxl")
#library("readxl")
#library("rgdal")
#library("maps")
#library("ggthemes")
#library("gganimate")
#library("RColorBrewer")
#library("geomnet")
#library("gridExtra")
#library("knitr")
#library("forecast")
#library("xtable")
#library(readr)
#library(repmis)

##########################################################################
########################################/// DATA ///
##########################################################################
load(url("https://github.com/fousseynoubah/ssm/blob/master/ssm_data.RData?raw=true"))

```


# Intro

When I started writing this post, I was fairly confident it would be quick.
After all, I had all the materials I needed, a bunch of scripts I had spent months writing.
It was supposed to be a matter of copy-and-paste.
I was wrong.
You see, once again, I underestimated the power of the *monster* in me who wants to get it right all the time.
Don't get me wrong.
It's always a good thing for one to strive to be their best self, and to give the best of what they've got.
In my case, unfortunately, that line always seems to get further away as I get close to it (I can only imagine the head some of my friends would make when reading this).
This post was no exception.

I was still learning ggplot2 when this data science project came about.
If I'm honest, I would say that I was still in my honeymoon phase.
I simply couldn't get enough of ggplot2.
So, it only seemed natural to use it for the project.
On the [GitHub repository](https://github.com/fousseynoubah/ssm) I created, my main outputs were pdf files.
So, I had no reason to think beyond ggplot2.
Then, the idea for this blog came, and I realized that interactive visualization would be a better way to go.
I had to chose a package.
Three options were available to me: ggvis, plotly, shiny.
Actually, these three would not be mutually exclusive, making the strategy even harder.

I had a strong leaning towards ggvis, on which I had already taken few courses online.
I loved it.
It was simply the follow-up act to ggplot2.
Apart from few adjustments I had to be mindful of (trading `+` for `%>%`, `=` for `=~` or `:=`, `geom` for `layers`, etc.), I felt home.
Unfortunately, the package is currently dormant.
That deterred a bit.

So, I considered the second option: shiny.
I had dabbled in that one too.
But just dabbled.
I knew about the basics, but not enough to feel confident to use it for an entire project.
I still hadn't gotten the hang of the all the details about the user interface and the server.
With what I knew of ggplot2, just a little would've sufficed to make me pick shiny, but I didn't want to embark on a another long learning journey.
So, I passed.

Then, entered the third option: plotly.
That one caught my eye a long time ago, long before I reconnected with `R`, but I was intimidated.
I thought it would be too complicated to learn (just like with shiny...seeing a pattern of laziness in here?).
Also, I was getting a little bit comfortable with the little I had learned.
In my mind, I could've gone a lifetime without ever needing plotly.
One detail made me reconsider my position: ggplotly.
Realizing that I could create an object with ggplot2, pass it to ggplotly, and have a plotly object was - to put it like teenagers - dope (please don't tell me I'm outdated on that one already).
But, then, I reached the limits of the function.
I was not getting everything.
Some details were being lost (like the subtitles and the captions, for instance).

In the end, I found myself in a pickle.
With ggplot2, I was going to miss out on the joys of interactive visualization.
With ggplot2 + plotly, I was going to lose some of what makes ggplot2 great.
And, with shiny + one of either, I was going to have to learn a lot.
I spent days weighing my options, going back and forth between chunks of code, hoping I would find a way around the issue.
I didn't.
But I knew one thing, though.
ggvis was not an option.
Unlike plotly or shiny, it was not being being maintained and developed.
For all I know, it could be in its grave (I dread that fate so much that I shiver at the thought of it), or taking a long nap before a comeback (like a Mommy...well a good one...like those released in the early 2000's...not the one we recently got from Tom Cruise).
I ended up spending long hours watching Youtube videos (which is now the place I go to learn rather than to watch latest music videos like I used to few years ago...looks like I'm growing up) and surfing websites with tutorials on plotly and shiny in R.
I still am not there yet, the place where I feel the confidence that only mastery can bring.
But I understood few tricks.
Mostly, I only set out *translate* my old ggplot2 graphs into interactive shiny-powered plotly ones.

What I took (or I'm still taking) away from the experience behind this post is this: there's not shortcut to learning and nothing is truly ever lost (waouw, I can be profound sometimes. Ok, mark that, and write it on my grave).
This project probably brought me closer to all the tools aforementioned more than any previous tutorials I took online.
Probably because, I'm less concerned with reproducing what I see from the instructor that I am making sense of the data I have through visualization.

This post was initially meant to be a simple rework of scripts I had previously written.
It ended up being a shiny/plotly-centered episode in this incidents-focused season (if you were/are a Lost fan, you'd get it).

# The incidents

<iframe width="700" height="600" scrolling="no" frameborder="no"  src="https://fousseynoubah.shinyapps.io/ssm1/"> </iframe>


<iframe width="700" height="600" scrolling="no" frameborder="no"  src="https://fousseynoubah.shinyapps.io/ssm2/"> </iframe>




```{r incidents_apercu_gph1, eval=FALSE, echo=FALSE, fig.align='center', out.width="90%"}

### Graph: Number of incidents, per month
incidents %>%
  mutate(n = 1) %>%
  group_by(mois, annee) %>%
  summarise(incidents = sum(n), victimes = sum(nbr_victimes), morts = sum(nbr_morts)) %>%
  mutate(jour = paste0(0, 1)) %>%
  mutate(date = paste(annee, mois, jour, sep = "-")) %>%
  mutate(date = ymd(date)) %>%
  select(annee, mois, jour, date, incidents, victimes, morts) %>%
  plot_ly(x = ~date, y = ~incidents, 
          type = "bar",
          marker = list(color = "darkred")) %>% 
  layout(title = "Number of incidents, per month",
         yaxis = list(title = "Incidents"),
         xaxis = list(title = "Date"))

```






```{r incidents_apercu_gph2, eval=FALSE, echo=FALSE, message=FALSE, warning=FALSE, fig.align='center', out.width="90%"}

### Graphique: nombre d'incidents par par mois, cumul
#incidents %>%
#  mutate(mois = factor(mois,
#                       levels = c(1:12),
#                       labels = c("Jan", "Feb", "Mar", "Apr", 
#                                  "May", "Jun", "Jul", "Aug", 
#                                  "Sep", "Oct", "Nov", "Dec"),
#                       ordered = TRUE
#                       )) %>% 
#  group_by(annee, mois) %>% 
#  summarise(incidents = n()) %>% 
#  mutate(cumsum = cumsum(incidents)) %>% 
#  select(-incidents) %>% 
#  spread(key = annee, value = cumsum) %>% 
#  plot_ly(x =~mois) %>% 
#  add_lines(y =~`2015`, name = '2015', mode = 'lines') %>% 
#  add_lines(y =~`2016`, name = '2016', mode = 'lines') %>% 
#  add_lines(y =~`2017`, name = '2017', mode = 'lines') %>% 
#  add_lines(y =~`2018`, name = '2018', mode = 'lines') %>% 
#  layout(title = "Cumulated number of incidents, per year",
#         yaxis = list(title = "Incidents"),
#         xaxis = list(title = "Month"),
#         annotations = list(x = 0.75, y = -0.15, 
#                            text = "<b>Source:</b> Data from Malilink website", 
#                   showarrow = F, xref='paper', yref='paper', 
#                   xanchor='left', yanchor='right', # xshift=0, yshift=0,
#                   font=list(size=10))# , margin = list(b = 50)
#         )

incidents %>%
  select(year = annee, month = mois, day = jour, date) %>% 
  mutate(week = week(date)) %>%
  group_by(year, week) %>% 
  summarise(incidents = n()) %>% 
  mutate(cumsum = cumsum(incidents)) %>% 
  plot_ly(x =~ week, y =~cumsum, color =~factor(year), mode = "lines") %>% 
  layout(title = "Cumulated number of incidents, per year",
       yaxis = list(title = "Incidents"),
         xaxis = list(title = "Week"),
         annotations = list(x = 0.75, y = -0.15, 
                            text = "<b>Source:</b> Data from Malilink website", 
                   showarrow = F, xref='paper', yref='paper', 
                   xanchor='left', yanchor='right', # xshift=0, yshift=0,
                   font=list(size=10))# , margin = list(b = 50)
         )


```


































